An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
reformulated some of the text about generalised Pa images

Note that while Parikh's theorem applies to context-free languages, in this
paper we are only interested in regular languages.

\subsection{Generalised Parikh maps}\label{sec:generalised}

To generalise the logic of the Parikh map, we introduce a morphism $\Map:\: \Sigma^*
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$
\end{itemize}

@@

Since Parikh images are semilinear, any Parikh image can be written as a set of linear
equations. The following construction, here adapted to work on an NFA $\Automaton = \AutomatonTuple$, was presented in\cite{generate-parikh-image}:
\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \longrightarrow  \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} \Filter(\Transition) - \sum_{\Transition = \FromLabelTo{}{}{\State}} \Filter(\Transition)
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{\State'}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \longrightarrow \StateVar_{\State} > 0
} \\
& \AndComp{\State \in F}{
  \StateVar_{\State} > 0 \longrightarrow \FinalStateVar_{\State} = 1
} \\
&\AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \longrightarrow
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates \land \right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

@@

\section{Applications of generalised Parikh images}\label{sec:applications}

\subsubsection{Example~1: String length}
\Fudge{We have already illustrated refer intro}
A useful example of such a simplifying morphism can express string length, the
problem that originally motivated our study of the Parikh map. We let $\Monoid =
\left(\mathbb{Z};+;0\right)$ and let $\Map(x) = 1$. Then the length of a string
$s = s_1 \RepeatSum{\Concat}  s_n$ is given by $\Map(s) = \sum \Map(s_i) = 1
\RepeatSum{+} 1$. Note that if we substitute vectors and vector addition for
this monoid, we obtain the regular Parikh map.

@@

\subsubsection{Example~2: Parikh Automata}\label{sec:parikh-automata}
\contents{\item Ostrich+!}

>>>>

Note that while Parikh's theorem applies to context-free languages, in this
paper we focus only on regular languages.

@@

Since Parikh images are semilinear, any Parikh image can be written as a set of linear
equations. The following construction, here adapted to work on an NFA $\Automaton = \AutomatonTuple$, was presented in \cite{generate-parikh-image}:
\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \longrightarrow  \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} \Filter(\Transition) - \sum_{\Transition = \FromLabelTo{}{}{\State}} \Filter(\Transition)
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{\State'}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \longrightarrow \StateVar_{\State} > 0
} \\
& \AndComp{\State \in F}{
  \StateVar_{\State} > 0 \longrightarrow \FinalStateVar_{\State} = 1
} \\
&\AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \longrightarrow
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates \land \right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

@@

\section{Generalised Parikh Maps}\label{sec:generalised}

It is easy to see that the Parikh map~$\ParikhMap$ represents a
homomorphism from the (free) non-commutative monoid~$\Strings$ to the
(free) commutative monoid~$\Naturals^k$. As we are often not
interested in the full Parikh image, but only in some projections of
it, for the purposes of this paper we use a generalised version of the
Parikh map. We consider \emph{arbitrary} homomorphisms
$\Map:\: \Sigma^* \to \Monoid$, where
$\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative
monoid; this means that $\Map(\epsilon) = 0_{\Monoid}$ and
$\Map(u \Concat w) = \Map(u) +_{\Monoid} \Map(w)$. We give several
examples of such generalised Parikh image later in this section.

Observe that every homomorphism~$\Map:\: \Sigma^* \to \Monoid$ can be
represented as the composition~$h' \circ \ParikhMap$, for some
homomorphism~$h' : \Naturals^k \to M$. One of the insights underlying our
approach is that it is more efficient to directly compute a
generalised Parikh image~$\Map(\Language)$, than to first compute the
standard image~$\ParikhMap(\Language)$ followed by projection to
some property of interest.

%\subsection{Applications of generalised Parikh images}\label{sec:applications}

\subsection{Example~1: Reasoning about string length}

A useful example of such a simplifying morphism can express string
length, the problem that originally motivated our study of the Parikh
map. We let $\Monoid = \left(\mathbb{Z};+;0\right)$, and define the
homomorphism~$\Map$ by $\Map(a) = 1$ for all
characters~$a \in \Alphabet$. Then the length of a string
$s = s_1 \RepeatSum{\Concat} s_n$ is given by
$\Map(s) = \sum \Map(s_i) = 1 \RepeatSum{+} 1 = n$.

@@

\subsection{Example~2: Parikh Automata}\label{sec:parikh-automata}
\contents{\item Ostrich+!}
